Transition path sampling and the calculation of free energies of enzymatic reactions

### 2022-03-31 tex/manuscript.tex
modified the introduction a little more

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcb,manuscript=article,reprint]{achemso}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{chemformula} % Formula subscripts using \ch{}
\usepackage[T1]{fontenc} % Use modern font encodings
\usepackage{graphicx}
\usepackage{caption}
\usepackage{amsmath}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We discuss a scheme for the calculation of free energies within 
  the transition path sampling (TPS) method. TPS is often used for
  the calculations of reaction coordinates of complex reactions in 
  the condensed phase and in this work we use it to sample the 
  equilibrium distribution of the reaction coordinate. This 
  distribution can then be used for the calculations of free energies
  using standard Boltzmann inversion schemes. We calculate the free
  energy profiles of the human Mat2a enzyme and the plasmodium 
  adenosine deaminase to demonstrate the applicability of this method
  for enzyme catalysis reactions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Atomistic level detailed study of the mechanism of enzyme catalysis 
reactions is necessary to understand their efficiency as well as to 
design molecules that can act as inhibitors for their activity. 
Molecular dynamics (MD) simulations 2 have often been used to study 
large biomolecular systems in atomistic detail since they provide an 
excellent computational cost to detail ratio. MD simulations to study 
chemical reactions that go through free energy barriers are often 
impractical to simulate since it would take a long time for the system 
to cross over barriers.
To overcome these energy barriers and 
to obtain ergodic sampling several methods have been developed over the years such as 
umbrella sampling, \cite{Kastner11WileyInterdiscipRevComputMolSci1p932} 
metadynamics, \cite{Barducci11WileyInterdiscipRevComputMolSci1p826}
steered MD, \cite{Park04JChemPhys120p5946}
milestoning, etc. These methods rely on applying bias
potentials for the system to overcome energy barriers and hence require the knowledge of 
the reaction coordinates.  
Transition path sampling \cite{Bolhuis02AnnRevPhysChem53p291,Dellago98JChemPhys108p1964}
(TPS) is an enhanced sampling method for 
rare events that transition a system between two stable states often 
referred to as reactants and products. 6

@@

Human Mat2a enzyme and plasmodium vivax adenosine deaminase \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618} 
\section{Theory}
\subsection{Transition path sampling}
%Transition path sampling is an importance sampling in the trajectory space.
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

\section{Adenosine deaminase reaction coordinates and free energy}
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.25]{./figures/ada-protein.png}
\end{figure}

\begin{figure}[ht]
\centering
\begin{minipage}[b]{0.32\linewidth}
\includegraphics[scale=0.3]{figures/reactant-state.png}
\caption{Reactant state}
\label{fig:minipage1}
\end{minipage}
\quad
\begin{minipage}[b]{0.32\linewidth}
\includegraphics[scale=0.3]{figures/int1.png}
\caption{Intermediate state}
\label{fig:minipage2}
\end{minipage}
\quad
\begin{minipage}[b]{0.32\linewidth}
\includegraphics[scale=0.3]{figures/product-state.png}
\caption{Product state}
\label{fig:minipage1}
\end{minipage}
\end{figure}

%\subsection{Outline}
\begin{figure}[ht]
\centering
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[scale=0.5]{figures/reac-sc-oc-167.png}
\caption{Trajectory 167}
\label{fig:minipage1}
\end{minipage}
\quad
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[scale=0.9]{figures/den-fenergy-1.pdf}
\caption{Free energy profile for the enzymatic reaction in human Mat2a enzyme}
\label{fig:minipage2}
\end{minipage}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jpcbfk,manuscript=article,layout=traditional]{achemso}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{chemformula} % Formula subscripts using \ch{}
\usepackage[T1]{fontenc} % Use modern font encodings
\usepackage{graphicx}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{mathptmx}
\usepackage[scaled=0.92]{helvet}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  Transition path sampling (TPS) has been widely applied for the
  calculations of transition state and reaction coordinates of 
  condensed phase systems. Here, we discuss a scheme for the calculation
  of free energies with the TPS method 
  We discuss a scheme for the calculation of free energies within 
  the transition path sampling (TPS) method. TPS is often used for
  the calculations of reaction coordinates of complex reactions in 
  the condensed phase and in this work we use it to sample the 
  equilibrium distribution of the reaction coordinate. This 
  distribution can then be used for the calculations of free energies
  using standard Boltzmann inversion schemes. We calculate the free
  energy profiles of the human Mat2a enzyme and the plasmodium 
  adenosine deaminase to demonstrate the applicability of this method
  for enzyme catalysis reactions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Molecular dynamics (MD) simulations are being 
used to calculate quantitative estimations of experimentally measurable 
kinetic and thermodynamic parameters of enzyme catalyzed reactions. 
\cite{Karplus02NatStructMolBiol9p646,
Schramm98AnnuRevBiochem67p693,Zhang05AccChemRes38p379,
Schramm11AnnuRevBiochem80p703,Schramm18ChemRev118p11194} 
This has helped in proposing mechanisms and explanations for 
enzyme activity, high throughput screening of millions of inhibitor 
molecues in drug discovery,

This is quite useful in explaining 
the efficiency and rate of such reactions, developing 
inhibitors for various enzymes in pharmacology, de novo design of 
enzymatic activity.
MD simulations model chemical reactions by taking the system from a reactant state 
to a product state via a high energy transition state.
To overcome these energy barriers and 
to obtain ergodic sampling several methods have been developed over the years such as 
umbrella sampling, \cite{Kastner11WileyInterdiscipRevComputMolSci1p932} 
metadynamics, \cite{Barducci11WileyInterdiscipRevComputMolSci1p826}
steered MD, \cite{Park04JChemPhys120p5946}
milestoning, etc. These methods rely on applying bias
potentials for the system to overcome energy barriers and hence require the knowledge of 
the reaction coordinates.  
Transition path sampling \cite{Bolhuis02AnnRevPhysChem53p291,Dellago98JChemPhys108p1964}
(TPS) is an enhanced sampling method for 
rare events that transition a system between two stable states often 
referred to as reactants and products. 6

@@

Human Mat2a enzyme and plasmodium vivax adenosine deaminase \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618} 
\section{Methods}
\subsection{Transition path sampling}
%Transition path sampling is an importance sampling in the trajectory space.
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

\section{Adenosine deaminase reaction coordinates and free energy}
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.22]{./figures/ada-protein.png}
\end{figure}

%\begin{figure}[ht]
%\centering
%\begin{minipage}[b]{0.32\linewidth}
%\includegraphics[scale=0.3]{figures/reactant-state.png}
%\caption{Reactant state}
%\label{fig:minipage1}
%\end{minipage}
%\quad
%\begin{minipage}[b]{0.32\linewidth}
%\includegraphics[scale=0.3]{figures/int1.png}
%\caption{Intermediate state}
%\label{fig:minipage2}
%\end{minipage}
%\quad
%\begin{minipage}[b]{0.32\linewidth}
%\includegraphics[scale=0.3]{figures/product-state.png}
%\caption{Product state}
%\label{fig:minipage1}
%\end{minipage}
%\end{figure}

%\subsection{Outline}
\begin{figure}[ht!]
\centering
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[scale=0.4]{figures/reac-sc-oc-167.png}
\caption{Trajectory 167}
\label{fig:minipage1}
\end{minipage}
\quad
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[scale=0.8]{figures/den-fenergy-1.pdf}
\caption{Free energy profile for the enzymatic reaction in human Mat2a enzyme}
\label{fig:minipage2}
\end{minipage}
\end{figure}
